How do changes in ex vary between countries?

### 2022-06-02 draft_working_paper.Rmd
try to tidy main analysis; build simple app to look at how changepoint models work

```{r, echo=FALSE}
# load packages
pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.
# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))
```

@@

```{r, echo = FALSE}
countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)
hmd_ex_selected_countries <- 
  hmd_lt %>% 
    filter(code %in% countries_of_interest) %>% 
    filter(Age %in% c(0, 65)) %>% 
    select(code, Year, Age, sex, ex) %>% 
    rename(x = Age, year = Year)
hmd_ex_selected_countries
# Code for producing synthetic Germany
source(here("R", "make_synthetic_germany_functions.R"))
series_east_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_f0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "female", 
  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
)
###
series_east_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_f65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "female", 
  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
)
##
series_east_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_m0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "male", 
  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
)
###
series_east_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_m65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "male", 
  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
)
hmd_ex_selected_countries_with_synth <- bind_rows(
  hmd_ex_selected_countries,
  tmp_df_m0,
  tmp_df_m65,
  tmp_df_f0,
  tmp_df_f65
)
# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )
```

@@

```{r, echo=FALSE}
change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

change_in_ex_selected_countries
```

@@

- [ ] make smoother lines coloured by sex
- [X] use consistent colours for all graphs showing separate sex results

@@

- [X] Give this a title and further info (lower priority)
- [X] Look at splitting by starting age 
    - [ ] Look at further refining starting age split

@@

### England and Wales

```{r, echo = FALSE, warning=FALSE, message = FALSE}

#ENW female at birth
f_e0_ENW <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_ENW <- f_e0_ENW %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ENW male at birth
m_e0_ENW <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

  m_e0_ENW <- m_e0_ENW %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #ENW female at age 65
f_e65_ENW <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable 
  f_e65_ENW <- f_e65_ENW %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ENW male at aged 65
m_e65_ENW <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

 m_e65_ENW <- m_e65_ENW %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

 #add labels
f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")

m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")

f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")

m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

  #to make grid, first assign to object
# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
# 
# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
# 
# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
# 
# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

#changing labels for grid to work
# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females")
# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", subtitle = "Life expectancy changes at birth for males")
# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65")
# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", caption = "Source: Human Mortality Database")
# #put side by side, needs package gridExtra then uses grid.arrange(plot1, plot2, ncol=2)
# grid.arrange(f_e0_ENW, m_e0_ENW, f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)
# #grid.arrange(f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)

#   hmd_ex_selected_countries_with_synth %>% 
#       filter(sex == "male") %>% 
#       filter(code %in% c("GBRTENW", "GBR_SCO")) %>% 
#     group_by(code, x) %>% arrange(year) %>% 
#       mutate(delta_ex = ex - lag(ex)) %>%
#     ungroup %>% 
#     filter(year >= 1979) %>% 
#     ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
#    geom_line() +
#    geom_hline(yintercept = 0) + 
#    stat_smooth(colour = "aquamarine4")+
# facet_grid(x ~ code)

  ex_ENW <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "GBRTENW") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "England and Wales", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")

### Scotland
```{r, echo = FALSE, warning=FALSE, message = FALSE}
#SCO female at birth
f_e0_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_SCO <- f_e0_SCO %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #SCO male at birth
m_e0_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

  m_e0_SCO <- m_e0_SCO %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #SCO female at age 65
f_e65_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_SCO <- f_e65_SCO %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #SCO male at aged 65
m_e65_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

 m_e65_SCO <-  m_e65_SCO %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4", show.legend = TRUE)

  #add labels
f_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 m_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
 f_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
 m_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

 ex_SCO <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "GBR_SCO") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_SCO + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

### France

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#FRA female at birth
f_e0_FRA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_FRA <- f_e0_FRA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #FRA male at birth
m_e0_FRA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

  m_e0_FRA <- m_e0_FRA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #FRA female at age 65
f_e65_FRA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_FRA <- f_e65_FRA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #FRA male at aged 65
m_e65_FRA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

 m_e65_FRA <-  m_e65_FRA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4", show.legend = TRUE)

  #add labels
f_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 m_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
 f_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
 m_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

 ex_FRA <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "FRATNP") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_FRA + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

### Spain

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#ESP female at birth
f_e0_ESP <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "ESP") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_ESP <- f_e0_ESP %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ESP male at birth
m_e0_ESP <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "ESP") %>% 
      filter(year >= 1979)

  m_e0_ESP <- m_e0_ESP %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #ESP female at age 65
f_e65_ESP <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "ESP") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_ESP <- f_e65_ESP %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ESP male at aged 65
m_e65_ESP <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "ESP") %>% 
      filter(year >= 1979)

 m_e65_ESP <-  m_e65_ESP %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4", show.legend = TRUE)

  #add labels
f_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 m_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
 f_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
 m_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

 ex_ESP <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "ESP") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_ESP + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

### Italy

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#ITA female at birth
f_e0_ITA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "ITA") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_ITA <- f_e0_ITA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ITA male at birth
m_e0_ITA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "ITA") %>% 
      filter(year >= 1979)

  m_e0_ITA <- m_e0_ITA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #ITA female at age 65
f_e65_ITA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "ITA") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_ITA <- f_e65_ITA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #ITA male at aged 65
m_e65_ITA <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "ITA") %>% 
      filter(year >= 1979)

 m_e65_ITA <-  m_e65_ITA %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   #Trying to add legend
   #stat_smooth(aes(colour = "aquamarine4"), show.legend = TRUE, labels = "Smoothed")
  stat_smooth(colour = "aquamarine4")

  #add labels
f_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 m_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
 f_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
 m_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

 ex_ITA <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "ITA") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_ITA + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

### Germany: Synthetic

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#DEUTS female at birth
f_e0_DEUTS <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "DEUTSYNTH") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_DEUTS <- f_e0_DEUTS %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #DEUTS male at birth
m_e0_DEUTS <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "DEUTSYNTH") %>% 
      filter(year >= 1979)

  m_e0_DEUTS <- m_e0_DEUTS %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #DEUTS female at age 65
f_e65_DEUTS <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "DEUTSYNTH") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_DEUTS <- f_e65_DEUTS %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #DEUTS male at aged 65
m_e65_DEUTS <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "DEUTSYNTH") %>% 
      filter(year >= 1979)

 m_e65_DEUTS <-  m_e65_DEUTS %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4", show.legend = TRUE)

  #add labels
f_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 m_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
 f_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
 m_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")

 #Grid
 ex_DEUTSYNTH <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "DEUTSYNTH") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_DEUTSYNTH + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

### Germany: East

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#DEUTE female at birth
f_e0_DEUTE <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "DEUTE") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e0_DEUTE <- f_e0_DEUTE %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #DEUTE male at birth
m_e0_DEUTE <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "DEUTE") %>% 
      filter(year >= 1979)

  m_e0_DEUTE <- m_e0_DEUTE %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4")

  #DEUTE female at age 65
f_e65_DEUTE <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "female") %>% 
      filter(code == "DEUTE") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
  f_e65_DEUTE <- f_e65_DEUTE %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "lightpink3")

  #DEUTE male at aged 65
m_e65_DEUTE <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 65) %>% 
      filter(sex == "male") %>% 
      filter(code == "DEUTE") %>% 
      filter(year >= 1979)

 m_e65_DEUTE <-  m_e65_DEUTE %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
   geom_line() +
   geom_hline(yintercept = 0) + 
   stat_smooth(colour = "aquamarine4", show.legend = TRUE)

>>>>

```{r, echo=FALSE}
# load packages
pacman::p_load(knitr, here, segmented, tidyverse, kableExtra)

# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)

source(here("R", "make_synthetic_germany_functions.R"))

@@

```{r, echo = FALSE}
source(here("R", "make_pop_selection.R"))
```

@@

```{r, echo=FALSE}
change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup() 
```

@@

@@

@@

```{r}
source(here("R", "annual_change_graphs_by_population.R"))
source(here("R", "breakpoint_functions.R"))
```
